Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-09-06 Manuscript/01-Intro.Rmd
Finish revise section 1

<!-- Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when Cleveland [@Cleveland1979] proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed. [@Wood2020] With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classic statistical models.  -->

@@

The proposed framework, BHAM differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB-GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting by using optimization algorithms instead of computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional models more feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent algorithm and EM-Iterative weighted least square algorithm for different utilities. The applications of the two aforementioned algorithm are fruitful in Bayesian hierarchical models in high-dimensional data analysis. [@Yi2012; @Rockova2014a; @Rockova2018] Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high dimensional setting. Furthermore, the proposed model addresses the incapability of bi-level selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), make transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical generalized additive model, introduce the proposed spike-and-slab spline priors, and describe the two fast-fitting algorithms. In Section \ref{sec:sim}, we compare the proposed framework to state-of-art models, mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies. Analyses of two metabolomics datasets using the proposed model are presented in Section \ref{sec:real_data}. Conclusion and discussions are given in Section \ref{sec:concl}.

<!-- Similar to expanding classic models to high-dimensional models, expanding the GAM framework for high-dimensional applications is not a trivial task. The difficulties mainly come in two-fold: grouping nature of the basis functions, and the balance between spline smoothness and variable sparsity. For GAM models, a smoothing function is consists of multiple basis functions. The basis functions of a variable is inherently related. The grouping structure should be properly taken care of for accurate estimation. Moreover, shrinkage is previously used solely for sparsity of the model in high dimensional regression models, or smoothness in GAMs. How to organically combine the two together remains as a troubling question.  -->

>>>>

<!-- Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when Cleveland [@Cleveland1979] proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed. [@Wood2020] With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classical statistical models.  -->

@@

The proposed framework, BHAM, differs from previous spike-and-slab based GAMs, i.e. the spike-and-slab GAM [@Scheipl2012] and the SB-GAM [@Bai2021] in many ways. First of all, the proposed spike-and-slab spline prior is a spike-and-slab lasso type prior using independent mixture double exponential distribution, compared to spike-and-slab GAM. Spike-and-slab lasso priors provide computational convenience during model fitting by using optimization algorithms instead of computational intensive sampling algorithms. They make fitting high- and ultrahigh-dimensional models more feasible without sacrificing performance in prediction and variable selection. Secondly, the optimization algorithms introduced in this article include EM-Coordinate Descent algorithm and EM-Iterative weighted least square algorithm for different utilities. The applications of the two aforementioned algorithm are fruitful in Bayesian hierarchical models in high-dimensional data analysis. [@Yi2012; @Rockova2014a; @Rockova2018] Even though SB-GAM with a group spike-and-slab lasso prior also uses EM-CD for model fitting, the proposed EM-IWLS provides uncertainty measures that could be further used for confidence bound calculation and hypothesis testing. EM-IWLS complements the lack of inference potential from EM-CD in a median and high dimensional setting. Furthermore, the proposed model addresses the incapability of bi-level selection in SB-GAM. Lastly, the R implementation of the Bayesian hierarchical additive model is available publicly via BHAM at [https://github.com/boyiguo1/BHAM](https://github.com/boyiguo1/BHAM), make transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical generalized additive model, introduce the proposed spike-and-slab spline priors, and describe the two fast-fitting algorithms. In Section \ref{sec:sim}, we compare the proposed framework to state-of-art models, mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies. Analyses of two metabolomics datasets are presented in Section \ref{sec:real_data}. Conclusion and discussions are given in Section \ref{sec:concl}.

<!-- Similar to expanding classical models to high-dimensional models, expanding the GAM framework for high-dimensional applications is not a trivial task. The difficulties mainly come in two-fold: grouping nature of the basis functions, and the balance between spline smoothness and variable sparsity. For GAM models, a smoothing function is consists of multiple basis functions. The basis functions of a variable is inherently related. The grouping structure should be properly taken care of for accurate estimation. Moreover, shrinkage is previously used solely for sparsity of the model in high dimensional regression models, or smoothness in GAMs. How to organically combine the two together remains as a troubling question.  -->
